---
title: "Some ggplot2 Graphs"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Introduction:  Movies dataset

This markdown file illustrates some basic graphs produced using the ggplot2 package.  We illustrate these graphs for some movie data.
The package ggplot2movies contains data on 28,819 movies from the Internet Movie Database (imdb.com).  Here are the variables:

- title. Title of the movie.

- year. Year of release.

- budget. Total budget (if known) in US dollars

- length. Length in minutes.

- rating. Average IMDB user rating.

- votes. Number of IMDB users who rated this movie.

- r1-10. Multiplying by ten gives percentile (to nearest 10%) of users who rated this movie a 1.

- mpaa. MPAA rating.

- action, animation, comedy, drama, documentary, romance, short. Binary variables representing if movie was classified as belonging to that genre.

I load this package and the tidyverse suite of packages that includes ggplot2.  We focus on a sample of 1000 of these movies that is put in the data frame movies_sample.

```{r}
library(tidyverse)
library(ggplot2movies)
set.seed(1234)
movies_sample <- sample_n(movies, 1000)
```

# One categorical variable

We use the cut function to creat a new categorical variable era based on the year variable.

```{r}
movies_sample %>% mutate(era = cut(year, 
          breaks = c(1890, 1959, 1979, 1999, 2006),
          labels = c("before 1960", "60s-70s", "80s-90s", 
                     "2000 or later"))) ->
  movies_sample
```

A bar graph of era is constructed using the geom_bar geom.  In this sample, the largest number of movies were made in the 80's-90's.

```{r}
ggplot(movies_sample) +
  geom_bar(mapping = aes(era))
```

# One numeric variable

How long are these movies?  We construct a histogram of the lengths (minutes) of these movies using the geom_histogram geom.  Interestingly, we see two clusters -- a large cluster centered about 90 minutes and a smaller cluster of shorter movies.

```{r}
ggplot(movies_sample) +
  geom_histogram(mapping = aes(length))
```

An alternative geom, geom_density, constructs and displays a density estimate of the movie lengths.

```{r}
ggplot(movies_sample) +
  geom_density(mapping = aes(length))
```

# Relationship between one categorical and one numeric variables

Has the length of the movie changed acrossed eras?  One can construct parallel boxplots where x (the categorical variable) is era and y (the response) is movie length.   On the average, movies seem to be a little longer, on average, in the 60s-70s and 80s-90s.

```{r}
ggplot(movies_sample) +
  geom_boxplot(mapping = aes(era, length))
```

An alternative way to graphically compare groups is by plotting points -- the geom_jitter function is used to avoid the overplotting of points.  This graph shows that there are many short movies of lengths 10 minutes or less.


```{r}
ggplot(movies_sample) +
  geom_jitter(mapping = aes(era, length))
```

# Relationship between two numeric variables

How has the average IMDB movie rating changed across years?  A scatterplot of year and rating is constructed by the geom_point geometric object.  The general pattern is assessed by drawing a lowess smooth on top of the scatterplot.  It is interesting that the older movies between 1930 and 1950 are rated higher than the movies between 1970 and 1990.

```{r}
ggplot(movies_sample) +
  geom_point(mapping = aes(year, rating)) +
  geom_smooth(mapping = aes(year, rating), method = "loess")
```

